
---
title: "Economic Growth and Health System Factors in Global Child Mortality"
author: "Collins Audi"
date: "2025-12-04"
output:
  html_document: default
  pdf_document: default
execute:
  warning: false
  message: false
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**BACKGROUND AND PURPOSE OF THIS HEALTH DATA ANALYSIS PROJECT**

Reducing the number of children who die before their fifth birthday is a core target within the sustainable development goal three of health and well-being and is a sign of a country’s human development index. Previously, the Preston curve showed a strong correlation between a country’s per capita income and under-5 deaths (Preston, 1975): whereby higher income levels are associated with lower under-five mortality.  Globally, the under-5 mortality has reduced by more than half from 94 per 1000 live births in 1990 to 37 per 1000 live births in 2023 (World Bank, 2023). This reduction can be attributed to economic growth and global health programs. 

Research in early 2000s established that economic growth alone is not enough to protect children unless it is paired with strategic priority-based investments in healthcare infrastructure. Houweling et al. (2005) reported that countries with comparable income levels experienced different outcomes because economic growth increased within-country inequalities and suggested increased public spending on health services like immunization and skilled birth attendance focusing on poor communities to close the gap. These inequalities were also confirmed by Cesar G Victoria (2003) who observed that the disparity between the rich and poor countries was becoming wider due to increasing income disparities within countries, and suggested increased interventions targeting the poor and equitable universal health coverage. 

Conversely, present evidence suggests that preventing mortality has evolved from access to the quality of care. Magret Kruk et al (2018) argues that despite the increased domestic and donor funding of health programs which has saved lives, the quality of care across different countries is often inadequate and recommends Universal Healthcare Coverage as an opportunity to improve the quality in health systems. The WHO Monitoring Report 2023 reported that while service coverage access improved across income status between 2015 and 2019, it was accompanied by limitations.According to this report, the service coverage index for healthcare has stagnated since the sustainable development goals were launched in 2015.The report further reveals how the focus of many countries has been on vertical programs rather than investing in health system capacities. In the report, while the low income and low middle income countries experienced catastrophic out-of-pocket expenditures despite service access gains, the upper middle income countries had gaps in access to services. This suggests that expenditure on healthcare by a country does not guarantee positive outcomes because it depends on the structural stability of the systems, yet this is not well quantified and documented. The report also revealed gaps in access to reproductive, maternal, neonatal, child and adolescent health services globally.While the importance of system pillars and spending is known, it remains unclear whether the efficiency of financial investment is conditional on the pre-existing structural maturity of the health system.

This project investigates these determinants of child survival through a quantitative analysis of cross national data, guided by two core research questions:

**Question**: To what extent do national income, government health spending, and preventive/infrastructural factors explain variations in child mortality, and does the effect of health spending vary by a country's income status, used as a proxy for structural health system maturity?

While high income status of countries may not automatically guarantee better access to healthcare as indicated above, it does have better health system stability at baseline. Hence, income status is utilized in this study as a proxy for health system stability and horizontal infrastructure required to for sustainable survival outcomes.
   
**Variables per country (2000-2002)**
    •	Under_5 mortality rate 
    •	Access to water and sanitation 
    •	Healthcare investment (as a % of GDP)
    •	GDP Per Capita (adjusted by Our World in Data to reflect living costs and inflation)
    •	Under-5 population size

  **Sources of Data**
     •	Our World in Data
     •	UNICEF
     •	World Bank
     
**STEP 1: IMPORTING, CLEANING AND MANIPULATING THE FIRST DATASET: GLOBAL UNDER_5 MORTALITY DATA**

**Step 1-Phase 1: Preparing the Mortality Data**
My first task was to take the raw mortality records from UNICEF and transform them into a format that works for statistical analysis. I followed these four steps:

**i. Importing the Data**: I loaded the UNICEF under 5 mortality file and performed an initial clean to remove any hidden spaces in the text that could cause errors later.

**ii. Selecting the Best Estimates**: The raw file contains three different numbers for every country: a high, low, and median estimate. I filtered the data to keep only the Median rows. This ensures my analysis is based on the most likely figures while removing unnecessary noise.

***iii. Cleaning Column Names***: When R imports data, it often adds an "X" in front of years (like X2000). I stripped these "X" prefixes away so the year columns would be clean and easy to read.

***iv. Reshaping for Analysis (The "Pivot")***: Most importantly, I changed the data from a "wide" layout to a "long" layout. Instead of having 30 separate columns for different years, I collapsed them into two columns: Year and Mortality Rate.

**v. The Result**: By the end of this step, I had a clean, vertical timeline for every country. This consistent structure is what allows me to merge the mortality data with other factors like wealth and health spending for the next part of my story.

```{r}
#Importing and managing the first data set by UNICEF: Under 5 mortality data
under_5_mortality <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/under_five_mortality.csv", header = TRUE, sep =',')

library(dplyr)
library(tidyr)
library(stringr)

colnames(under_5_mortality) <- str_trim(colnames(under_5_mortality))   # remove leading/trailing spaces

under_5_mortality <- under_5_mortality[, 1:37] %>%
  filter(bound == "Median") %>%   # keep only Median rows
  rename_with(~ gsub("^X", "", .x), 5:37)   # remove 'X' prefix from columns 5 to 37

# Step 1: Keeping only Median rows necessary, leaving upper and lower boundaries since our interest is in modelling.

colnames(under_5_mortality)

# Step 2: Pivot year columns into long format
under_5_mortality <- under_5_mortality %>%
  pivot_longer(
    cols = 5:37,
    names_to = "year",
    values_to = "mortality_r",
  ) %>%
  mutate(year        =  as.integer(year),
         mortality_r =  round(mortality_r, 1))


```
**Step 1-Phase 2: Importing the Health and Economic Indicators**

To build a complete picture of the factors influencing under 5 mortality, I imported and processed five additional datasets. I followed these steps for each one:

**i. Importing the Key Variables**: I loaded datasets for Vaccine Coverage (DPT shots), Healthcare Expenditure (as a percentage of GDP), GDP per capita, WASH Access (clean water and sanitation), and the total Under 5 Population.

**ii. Standardizing the Values**: To ensure the data was precise but easy to read, I applied a rounding function to all numeric columns. I rounded the percentages for vaccines and WASH, as well as the dollar amounts for GDP, to one decimal place.

**iii. Data Alignment**: Each of these files was checked to ensure they contained the necessary country and year information. This allows these specific health and economic indicators to be matched perfectly with the mortality data I prepared in Phase 1.

**The Result**: By the end of this phase, I had a collection of clean, standardized tables. Each table represents a different "piece of the puzzle"—from a country's total wealth and population size to its specific investments in health and infrastructure. These are now ready to be combined into a single master dataset.
```{r}
#Importing and managing the immunization data set
immunisation <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/immunisation.csv", header = TRUE, sep =',')

#Rounding off immunization coverage data to one decimal place
immunisation  <- immunisation %>%
  mutate(dpt_coverage = round(dpt_coverage, 1))

#Importing and managing the healthcare expenditure data set as a % of GDP
h_expenditure <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/health_exp.csv", header = TRUE, sep =',')

#Rounding off health expenditure per capita figures to one decimal place
h_expenditure <- h_expenditure %>%
  mutate(health_exp_perc_ppp = round(health_exp_perc_ppp, 1))

#Importing gdp per capita data from Our World in Data
gdp_per_capita <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/gdp.csv", header = TRUE, sep =',')

#Rounding off gdp per capita to one decimal place
gdp_per_capita <- gdp_per_capita %>%
  mutate(gdp_percapita = round(gdp_percapita, 1))

#Importing and managing the access to WASH (access to clean water) data set
wash <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/wash.csv", header = TRUE, sep =',')

wash <- wash %>%
  mutate(
    wash_access = round(wash_access, 1))

#Importing Under 5 population data set
under5_pop <- read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/population.csv", header = TRUE, sep =',')


#Importing country income status data set
income <- read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/income_country.csv", header = TRUE, sep =',')

```
**Step 1-Phase 3: Data Merging and Coverage Audit**

In this final processing phase, I combined all individual datasets into one master file and conducted a thorough audit to check the quality of our timeline across different countries.

**Merging into a Master Dataset**: Using a "left join" function, I connected the mortality data with the five other datasets (Vaccines, WASH, GDP, Population, and Health Spending). I used the unique Country Code and Year as the keys to ensure every piece of information aligned perfectly across all sources.

**Cleaning the Combined Table**: After merging, the dataset contained several duplicate country name columns. I removed these extra columns and filtered the data to ensure that every row included the three most important variables for our bubble chart in the next exercise: GDP per capita, Mortality Rate, and Under 5 Population.

**Auditing the Timeline**: I conducted an audit to see how consistent the timeline was across different nations. I compared every country against the expected 23 year span (2000 to 2022).

**Full Range Coverage**: The audit revealed that 177 countries have a complete, uninterrupted record for the entire 23 year period.

**Partial Coverage**: Only 9 countries were identified as having partial or missing years. I kept these in the dataset to maintain global diversity, as reporting gaps are common in fragile or developing states.

**Final Export**: Once the data was verified, I exported the final collection as "merged_data.csv." This master file now contains 186 countries in total, providing a robust mix of stable and developing nations.

**The Result**: I now have a high quality master dataset. The fact that 177 countries have a full data range means that the patterns we see in the bubble chart are backed by a very large and reliable body of global evidence. This structure allows us to see not just where countries are today, but how they have moved over time.
```{r}
#Merging the datasets into a master dataset using iso_code and year columns, enabled by the left_join function
merged_df <- under_5_mortality %>%
  left_join(immunisation,  by = c("iso_code", "year")) %>%
  left_join(wash,          by = c("iso_code", "year")) %>%
  left_join(gdp_per_capita,by = c("iso_code", "year")) %>%
  left_join(under5_pop,    by = c("iso_code", "year")) %>%
  left_join(h_expenditure, by = c("iso_code", "year")) 


merged_df <- merged_df %>%
  select(
    -country.y,
    -country.y.y,
    -country.y.y.y,
    -country.x.x,
    -country.x.x.x,
    -bound,
     country = country.x) %>% 
  filter(!is.na(gdp_percapita), 
         !is.na(mortality_r), 
         !is.na(pop_under_5))

sapply(merged_df, class)

write.csv(merged_df, "merged_data.csv", row.names = FALSE)

#Now, we want to see the number of countries which have different year ranges 2000-2022 and 2001-2022
#Step 1: Counting the years available per country
country_years <- merged_df %>%
  group_by(iso_code, country) %>% # 1️⃣ Group the data by country
  summarise(
    min_year = min(year),         # 2️⃣ Earliest year for that country
    max_year = max(year),         # 3️⃣ Latest year for that country
    n_years  = n_distinct(year)   # 4️⃣ Count of unique years
  ) %>%
  ungroup() 
#group_by(iso_code, country.x) → ensures calculations are done separately for each country.
#min(year) → finds the earliest year recorded for that country.
#max(year) → finds the latest year recorded.
#n_distinct(year) → counts how many unique years exist for that country.
#ungroup() → removes grouping so we can do further operations easily.

#Step 2: Defining the expected full year range
expected_years <- 2000:2022           # 1️⃣ Full range of years
n_expected <- length(expected_years)  # 2️⃣ How many years: 23

#Step 3: Checking which countries have full 2000–2022 coverage
country_years <- country_years %>%
  mutate(
    full_2000_2022 = (min_year == 2000 & max_year == 2022 & n_years == n_expected))

#mutate() → adds a new column
#full_2000_2022 → TRUE if all conditions met:
#Earliest year = 2000
#Latest year = 2022
#Total number of years = 23
#FALSE → country is missing some years

#Step 4: Counting how many countries have full or partial coverage
country_years %>%
  summarise(
    full_range = sum(full_2000_2022),         # 1️⃣ Count TRUEs → complete countries
    partial_or_missing = sum(!full_2000_2022) # 2️⃣ Count FALSEs → incomplete countries
  )
```
*PHASE 2: VISUALIZING THE RELATIONSHIP BETWEEN UNDER-5 MORTALITY AND GDP**

As a preliminary step for answering the first question: **To what extent is a country’s GDP per capita associated with its under-five mortality rate (per 1000 live births) and does the link get weaker as countries get richer?**

The goal of this phase was to create an interactive "Preston Curve" (bubble chart) to visually explore the link between economic growth and child survival from 2000 to 2022.

**Dynamic Motion**: I built an animated bubble chart using Plotly, allowing us to see how countries move across the economic spectrum over time.

**Dimensions of Data**: Each bubble represents a country, with its position showing GDP per capita (x-axis) and Mortality Rate (y-axis). The size of each bubble is weighted by the Under-5 Population, highlighting where the largest groups of vulnerable children are located.Regional 

**Context**: I applied a custom color palette to group countries by their SDG Regions, making it easy to identify geographic trends and outliers.

**Logarithmic Scaling**: Because the differences between the richest and poorest nations are so vast, I used log scales for both axes. This ensures that progress in low-income countries is just as visible as trends in high-income ones.

**Preliminary Findings**: In this preliminary visualization, the bubble chart hints at a significant negative correlation between a country’s wealth and its child survival outcomes, suggesting that higher GDP per capita is generally linked to lower under-five mortality rates. This relationship follows a clear logarithmic pattern known as the Preston Curve: in lower-income countries, small increases in wealth are associated with declines in mortality, whereas in wealthier nations, the link becomes significantly weaker as survival rates approach the minimum possible death rate. For example, in 2000, Bangladesh (GDP - $2,712.9) had a high mortality rate of 85.4 per 1,000, but by 2022, as its wealth grew, it saw a dramatic drop in deaths to 31 deaths per 1000 live births. Conversely, in high-income regions like Europe and Northern America, bubbles are clustered at the bottom-right of the chart, showing that even substantial further increases in GDP result in only marginal improvements to already low mortality rates. Furthermore, the varying bubble sizes representing the under-five population highlight that high-population, lower-income nations like India and Bangladesh carry a disproportionate share of the global mortality burden despite their steady economic progress

```{r}
# ==========================================
# 1. LOAD LIBRARIES
# ==========================================
library(dplyr)    
library(ggplot2)  
library(plotly)   
library(scales)   

# ==========================================
# 2. DATA CLEANING & PRE-PROCESSING
# ==========================================
merged_df$country <- iconv(merged_df$country, from = "", to = "UTF-8", sub = "")
merged_df$country <- trimws(merged_df$country)
merged_df$country <- as.character(merged_df$country)

merged_df <- merged_df %>%
  filter(!is.na(mortality_r), country != "", !is.na(iso_code))

anim_df <- merged_df %>% 
  filter(year >= 2000, year <= 2022) %>% 
  group_by(iso_code, year) %>%          
  slice(1) %>%                           
  ungroup() %>%                          
  arrange(iso_code, year)               

# Strictly 1K and 100K for GDP; 1 and 500 for Mortality
x_breaks <- c(1000, 100000)  
y_breaks <- c(1, 500)       

# ==========================================
# 3. GGPLOT BASE CONSTRUCTION
# ==========================================
# Harmonized SDG region colors (soft, dark-background-friendly)
sdg_colors <- c(
  "Europe and Northern America"      = "#11A7B1", # Cyan/Teal
  "Northern Africa and Western Asia" = "#8E7AE3", # Purple
  "Central and Southern Asia"        = "#FF66C4", # Pink
  "Eastern and South-Eastern Asia"   = "#00B4FF", # Light Blue
  "Latin America and the Caribbean"  = "#4ECDC4", # Turquoise
  "Oceania"                          = "#FFA07A", # Light Salmon
  "Sub-Saharan Africa"               = "#FF7F7F"  # Light Red
)

p <- ggplot(
  anim_df,
  aes(
    x = gdp_percapita,       
    y = mortality_r,         
    color = sdg_region,       # Color by SDG region
    size = pop_under_5,      
    frame = year,            
    ids = iso_code,          
    group = iso_code,        
    text = paste(            
      "<b>", country, "</b>",
      "<br>Region:", sdg_region,
      "<br>Year:", year,
      "<br>GDP per Capita:", dollar(gdp_percapita),
      "<br>Under-5 Mortality:", round(mortality_r, 1),
      "<br>Under-5 Population:", comma(pop_under_5)
    )
  )
) +
  geom_point(alpha = 0.65) +  # Slightly transparent to reduce shouting

  scale_x_log10(
    limits = c(700, 150000),         
    breaks = x_breaks,               
    labels = dollar,                 
    expand = expansion(mult = 0.05)  
  ) +

  scale_y_log10(
    limits = c(1, 500),              
    breaks = y_breaks,               
    labels = comma,                  
    expand = expansion(mult = 0.05)
  ) +

  scale_size(range = c(4, 25), guide = "none") +
  scale_color_manual(values = sdg_colors, name = "SDG Region") +

  labs(
    x = "GDP per capita (USD, PPP Adjusted)",
    y = "Under-5 mortality (per 1,000 live births)"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "black", color = NA),  
    panel.background = element_rect(fill = "black", color = NA),
    text = element_text(color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.margin = margin(25, 25, 25, 25),
    axis.line = element_line(color = "white", linewidth = 0.6),
    axis.ticks = element_line(color = "white"),
  )

# ==========================================
# 4. PLOTLY CONVERSION & INTERACTIVITY
# ==========================================
interactive_anim <- ggplotly(p, tooltip = "text") %>%
  animation_opts(
    frame = 4000,      
    transition = 4000, 
    easing = "cubic-in-out", 
    redraw = FALSE     
  ) %>%
  layout(
    xaxis = list(
      title = list(text = "GDP per capita (USD)", standoff = 25, font = list(size = 14, color = "white")),
      tickvals = log10(x_breaks), 
      ticktext = dollar(x_breaks),
      tickfont = list(size = 11, color = "white"),
      showgrid = FALSE
    ),
    yaxis = list(
      title = list(text = "Under-5 mortality (per 1,000 live births)", standoff = 15, font = list(size = 14, color = "white")),
      tickvals = log10(y_breaks),
      ticktext = comma(y_breaks),
      tickfont = list(size = 11, color = "white"),
      showgrid = FALSE
    ),
    title = list(
      text = paste0(
            # Line 1: GDP and outcome + "continued"
            "<span style='font-size:18px; color:#FFFFFF;'>While </span>",
            "<span style='font-size:18px; color:#11A7B1;'><b>higher GDP</b></span>",
            "<span style='font-size:18px; color:#FFFFFF;'> is generally associated with lower under-5 mortality, continued</span><br>",
            
            # Line 2: Key requirements
            "<span style='font-size:18px; color:#FFFFFF;'> progress requires </span>",
            "<span style='font-size:18px; color:#7CFC9F;'><b>sustainable health financing</b></span>",
            "<span style='font-size:18px; color:#FFFFFF;'> and reduced </span>",
            "<span style='font-size:18px; color:#FFFFFF;'><b>health inequities</b></span>."

      ),
      x = 0.5, xanchor = "center", y = 0.88, yanchor = "top",
      font = list(family = "Arial, sans-serif")
    ),
    annotations = list(
      list(
        x = 0.88, y = 0.99, xref = "paper", yref = "paper",
        text = paste0(
          "<span style='color:#87CEEB'>Bubble size = Under-5 population</span>",
          "<span style='color:#cccccc'> &nbsp;&nbsp;|&nbsp;&nbsp; </span>",
          "<span style='color:#FFA07A'>Each bubble = A country (hover for details)</span>"
        ),
        showarrow = FALSE,
        font = list(size = 12, family = "Arial, sans-serif")
      )
    ),
    margin = list(t = 130, b = 80, l = 100, r = 40),
    showlegend = FALSE
    )

# Display interactive plot
interactive_anim

# ==========================================
# 5. EXPORT
# ==========================================
htmlwidgets::saveWidget(
  interactive_anim,
  "under5_mortality_final.html",
  selfcontained = TRUE 
)

```



```{r}
library(tidyverse)

library(dplyr)

merged_df2 <- under_5_mortality %>%
  left_join(immunisation %>% select(-any_of("country")),   by = c("iso_code", "year")) %>%
  left_join(wash %>% select(-any_of("country")),           by = c("iso_code", "year")) %>%
  left_join(gdp_per_capita %>% select(-any_of("country")), by = c("iso_code", "year")) %>%
  left_join(under5_pop %>% select(-any_of("country")),     by = c("iso_code", "year")) %>%
  left_join(h_expenditure %>% select(-any_of("country")),  by = c("iso_code", "year")) %>%
  left_join(income %>% select(-any_of("country")),         by = c("iso_code", "year")) %>%
  # Drop bound if it exists
  select(-any_of("bound")) %>%
  # This removes a row ONLY if ALL THREE are missing at the same time
  filter(!(is.na(wash_access) & is.na(health_exp_perc_ppp) & is.na(gdp_percapita)))

# Convert country names to UTF-8 encoding to handle special characters
# iconv() converts character vectors between encodings
# 'from = ""' uses the current locale's encoding
# 'to = "UTF-8"' specifies target encoding
# 'sub = ""' replaces unconvertible characters with empty string
merged_df2$country <- iconv(merged_df2$country, from = "", to = "UTF-8", sub = "")

# Remove leading/trailing whitespace from country names
# trimws() eliminates spaces, tabs, newlines at the beginning/end of strings
merged_df2$country <- trimws(merged_df2$country)

# Convert country column to character type (ensures proper string handling)
# as.character() explicitly converts to string data type
merged_df2$country <- as.character(merged_df2$country)

# 1. Define your exact target variables
target_vars <- c("mortality_r", "dpt_coverage", "wash_access", 
                 "gdp_percapita", "pop_under_5", "health_exp_perc_ppp")

# 2. Generate the "Problem Only" Table
failing_variables_audit <- merged_df2 %>%
  # Group by country to check each one's data health
  group_by(iso_code, country) %>%
  summarise(
    across(all_of(target_vars), ~ mean(is.na(.)) * 100), # Calculate % missing
    .groups = "drop"
  ) %>%
  # Reshape to see each variable on its own row for every country
  pivot_longer(cols = all_of(target_vars), 
               names_to = "Indicator", 
               values_to = "Missing_Percentage") %>%
  # FILTER: Only show cases that are 50% or above
  filter(Missing_Percentage >= 50) %>%
  arrange(country, desc(Missing_Percentage))

# 3. View the Resulting List
# This will ONLY show you the countries and variables that "fail" the rule.
print(failing_variables_audit)

# 1. Identify countries that are 100% missing in any variable
# MICE cannot impute if there is zero data to start with
critical_failures <- failing_variables_audit %>%
  filter(Missing_Percentage == 100) %>%
  distinct(iso_code) %>%
  pull(iso_code)

# 2. Identify countries failing more than 2 variables
high_failure_countries <- failing_variables_audit %>%
  group_by(iso_code) %>%
  count() %>%
  filter(n > 2) %>%
  pull(iso_code)
high_failure_countries

# 3. Combine and remove
final_drop_list <- unique(c(critical_failures, high_failure_countries))

merged_df3 <- merged_df2 %>%
  filter(
    !iso_code %in% final_drop_list,
    year >= 2000 & year <= 2022
  )


#Checking the structure of the remaining data
cat("=== TIER 1 DATA (All Countries) ===\n")
cat("Dimensions:", dim(merged_df3), "\n")
cat("Countries:", n_distinct(merged_df3$country), "\n")
cat("Years range:", min(merged_df3$year), "-", max(merged_df3$year), "\n")
cat("Variables:", names(merged_df3), "\n")
```
**Checking Missingness Patterns**
Missingness pattern was checked in details using heat maps.

```{r}
library(VIM)

aggr(merged_df3, prop = FALSE, combined = TRUE, numbers = TRUE, 
     cex.axis = 0.7, varheight = FALSE, sortVars = TRUE)

# Create a 'Missingness Flag' for the heatmap
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Reshape the data to long format
long_missing <- merged_df3 %>%
  select(country, year, wash_access, health_exp_perc_ppp, gdp_percapita, dpt_coverage) %>%
  pivot_longer(cols = c(wash_access, health_exp_perc_ppp, gdp_percapita, dpt_coverage), 
               names_to = "variable", 
               values_to = "value") %>%
  mutate(is_missing = is.na(value))

# 2. Create the Heatmap
# We use 'facet_grid' to separate variables while keeping countries on the Y-axis
p <- ggplot(long_missing, aes(x = year, y = country, fill = is_missing)) +
  geom_tile(color = "white") +
  facet_wrap(~variable) +  # This creates 3 side-by-side heatmaps
  scale_fill_manual(values = c("grey90", "red"), 
                    labels = c("Data Present", "Missing (NA)")) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7),
        strip.text = element_text(face = "bold")) + 
  labs(title = "Detailed Missingness: Country vs. Year vs. Variable",
       x = "Year", y = "Country")
p
# 3. Save as a TALL image to ensure names are readable
ggsave("Detailed_Variable_Heatmap.png", plot = p, width = 15, height = 25, limitsize = FALSE)
```

**Checking if missingness is dependent on other variables**

Visual analysis revealed systematic missingness patterns across key variables:

**1. WASH Indicators:**

- Countries with missing WASH data show mortality rate distributions shifted toward substantially higher values (extending toward 300-500 range)
- Countries with observed WASH data show mortality concentrated at lower values (0-100 range)
**Pattern:** Missingness strongly associated with worse health outcomes

**2. Health Expenditure:**
- Different population distributions between observed and missing groups
- Pattern: Missingness relates to country characteristics, not random

**3. GDP Data:**
- Missing GDP associated with higher mortality distributions
- Missing GDP concentrated in earlier years (2000-2010)
**Pattern:** Dual association with both poor outcomes and historical reporting gaps

**Missingness Mechanism Assessment**
**MCAR (Missing Completely at Random):** Ruled out due to systematic patterns.
**MAR (Missing at Random):** Plausible if key predictors are included
**mortality_r** for WASH/GDP missingness
**pop_under_5** for health expenditure missingness year for GDP missingness

**MNAR (Missing Not at Random):** Probable for WASH/GDP given extreme mortality associations.

**Imputation Justification**
MICE is conditionally justified because:
- Missingness patterns are explainable by observed variables
- Including key predictors makes MAR assumption tenable
```{r}
library(mice)
library(lattice) # For vignette-style histograms
library(dplyr)

# Create indicators: TRUE = Missing, FALSE = Observed
merged_df3 <- merged_df3 %>%
  mutate(
    M_wash   = is.na(wash_access),
    M_gdp    = is.na(gdp_percapita),
    M_health = is.na(health_exp_perc_ppp),
    M_dpt    = is.na(dpt_coverage)
  )

# A. Check if Mortality predicts missingness in WASH (Target 1)
p1 <- histogram(~ mortality_r | M_wash, data = merged_df3,
          main = "Mortality Rate by WASH Missingness",
          strip = strip.custom(factor.levels = c("WASH Obs", "WASH Miss")))

# B. Check if Mortality predicts missingness in GDP (Target 2)
p2 <- histogram(~ mortality_r | M_gdp, data = merged_df3,
          main = "Mortality Rate by GDP Missingness",
          strip = strip.custom(factor.levels = c("GDP Obs", "GDP Miss")))

# C. Check if Pop Under 5 predicts missingness in Health Exp (Target 3)
p3 <- histogram(~ pop_under_5 | M_health, data = merged_df3,
          main = "Pop by Health Exp Missingness",
          strip = strip.custom(factor.levels = c("Health Obs", "Health Miss")))

p4 <- histogram(~ year | M_wash, data = merged_df3,
                main = "Year by WASH Missingness",
                strip = strip.custom(factor.levels = c("WASH Obs", "WASH Miss")))

library(gridExtra)

grid.arrange(p1, p2, p3, p4, ncol = 2)
```
**Initial Diagnosis** Missing data was identified as Missing At Random (MAR) because missingness patterns depended on observed variables like Year and Mortality Rate.

**Problem 1:** Impossible Values & Poor Shape: The initial model (2l.pan) produced impossible negative values and failed to replicate the "humps" in the data distribution.

**Change:** Switched to 2l.pmm (Predictive Mean Matching) from the miceadds package.

**Reason:** PMM ensures no negative results by "borrowing" real observed values, which preserves the natural bimodal shape of your data.

**Problem 2:** "Optimistic" Bias: Density plots showed that imputed values for wash_access were too high, missing the lower-access (0–50) hump entirely.

**Change:** Added sdg_region and income_status to the predictor matrix.

**Reason:** These variables provided the necessary economic and geographic "clues" for the model to correctly identify and match missing data to low-access donors.

**Problem 3:** Formula Syntax Errors: The code crashed with "unexpected symbol" errors.

**Change:** Used gsub to replace spaces/hyphens with underscores and converted strings to factors.

**Reason:** R multilevel formulas cannot process special characters or raw strings within factor levels.

**Problem 4:** Numerical Spikes & Scale Warnings: GDP data produced "spiky" distributions (over-smoothing) and scaling warnings due to its high numerical range.

**Change:** Increased iterations to maxit = 30 and integrated categorical anchors.

**Reason:** This stabilized the model, allowing for a diverse spread of imputed values that matched the original distribution despite the large scale.

**Final Result:** Validated the success via stripplots and density plots, confirming that all imputed values are now bounded, realistic, and capture the true bimodal nature of the variables.
```{r}
# ============================================================================
# REVISED MICE PIPELINE (UNSCALED GDP + INCOME STATUS)
# ============================================================================

#| warning: false
#| message: false

# 1. LOAD LIBRARIES ----------------------------------------------------------
library(mice)
library(miceadds)  # Essential for 2l.pmm
library(dplyr)
library(lattice)

# 2. PREPARE DATA ------------------------------------------------------------
df_prepared <- merged_df3 %>%
  filter(!(is.na(wash_access) & is.na(dpt_coverage) & is.na(gdp_percapita))) %>%
  mutate(
    iso_code = as.integer(factor(iso_code)), 
    # CLEAN CATEGORICALS: Critical for formula compatibility
    sdg_region = as.factor(gsub("[[:space:]|-]", "_", as.character(sdg_region))),
    income_status = as.factor(gsub("[[:space:]|-]", "_", as.character(income_status))),
    year = as.numeric(year),
    time_in_country = year - 2000
  )

# 3. INITIALIZE MICE ---------------------------------------------------------
init <- mice(df_prepared, maxit = 0, printFlag = FALSE)
meth <- init$method
pred <- init$predictorMatrix

# 4. IMPUTATION METHODS ------------------------------------------------------
# 2l.pmm prevents negative predictions naturally
meth["wash_access"] <- "2l.pmm"
meth["gdp_percapita"] <- "2l.pmm"
meth["health_exp_perc_ppp"] <- "2l.pmm"
meth["dpt_coverage"] <- ""        
meth["sdg_region"] <- "polyreg"
meth["income_status"] <- "polyreg" 

# 5. PREDICTOR MATRIX --------------------------------------------------------
pred <- make.predictorMatrix(df_prepared)
pred[, "country"] <- 0

# Using raw gdp_percapita as requested
essential_preds <- c("mortality_r", "pop_under_5", "time_in_country", 
                     "sdg_region", "income_status", "gdp_percapita")

for (var in c("wash_access", "gdp_percapita", "health_exp_perc_ppp")) {
  pred[var, ] <- 0
  pred[var, essential_preds] <- 1
  pred[var, "iso_code"] <- -2     # Set iso_code as the Level-2 cluster
  pred[var, "dpt_coverage"] <- 1
}

# 6. RUN IMPUTATION ----------------------------------------------------------
set.seed(123)
imp_complete <- mice(
  df_prepared,
  m = 10,
  method = meth,
  predictorMatrix = pred,
  maxit = 30, # Increased iterations for better convergence with large scales
  printFlag = FALSE
)

# 7. VALIDATION -------------------------------------------------------------
# Note: Step 9 (pmax/pmin) is deleted. PMM handles bounds.
densityplot(
  imp_complete,
  ~ wash_access + gdp_percapita + health_exp_perc_ppp,
  layout = c(3, 1),
  main = "Final Validation: Observed vs Imputed"
)

# 8. CONVERGENCE (TRACE) PLOTS -----------------------------------------------
# This creates plots for the mean and SD of each variable across iterations
plot(imp_complete, 
     c("wash_access", "gdp_percapita", "health_exp_perc_ppp"),
     main = "Convergence Check: Mean and SD per Iteration")

# 9. STRIP PLOTS (Another way to see Imputation numbers) ----------------------
# This plots the raw imputed values side-by-side for each imputation dataset
stripplot(
  imp_complete, 
  wash_access + gdp_percapita + health_exp_perc_ppp ~ .imp, 
  pch = 20, 
  cex = 0.5,
  main = "Individual Imputed Values by Imputation Number"
)

# View the first imputed dataset (m = 1)
data_final <- complete(imp_complete, 1)
write.csv(data_final, "final_imputed.csv")
# 10. SAVE OUTPUT ------------------------------------------------------------
saveRDS(imp_complete, "imputed_data_unscaled.rds")
```
**Regression Analysis of the Final Panel Data with Imputed Values**
**Analysis Process and Model Selection**
The study utilizes a 22-year panel dataset of 122 countries to examine the determinants of under-five mortality. The analysis focuses on the relationship between child survival (ln\_mort) and its primary drivers: financial investment (ln_spend), infrastructure access (wash_access), and preventive healthcare (dpt_coverage). Income status serves as a proxy for structural maturity, representing the permanent foundations of a health system such as trained staff and stable supply chains. All financial and mortality variables were log-transformed to measure elasticities, showing the percentage change in survival for every 1% increase in spending.To ensure rigorous model selection, a multi-stage econometric hierarchy was followed. First, an F-test confirmed the necessity of accounting for country-specific characteristics. Second, a Hausman Test yielded a highly significant p-value (6.472e-06), rejecting Random Effects and confirming that fixed characteristics must be addressed. While the final estimation utilizes Median Quantile Regression (tau = 0.5) to compare effects across different development tiers, the inclusion of Income Status and the use of bootstrapped standard errors directly address the structural stability and autocorrelation identified in the diagnostic tests. This approach provides a reliable measure of the typical country experience while remaining resistant to outliers and heteroscedasticity.

**Key Findings**
In health economics, spending efficiency is defined as the capacity of a national health system to convert financial inputs into measurable biological outcomes, such as child survival. In the context of this model, efficiency is represented by the coefficients of the interaction terms. While the baseline coefficient for health spending (ln_spend) is negative, indicating that increased investment generally reduces mortality, the positive coefficients for the low-income interaction terms reveal an "efficiency penalty." This implies that in environments with lower baseline health system stability, financial resources face structural barriers that dilute their impact. Conversely, the negative baseline spending effect is most potent in high-income settings, where mature infrastructure allows money to be more effectively translated into lives saved.1. Integration with Global Vaccine and Infrastructure LiteratureThe median quantile regression results for the 2000–2022 period demonstrate that global reductions in child mortality are driven more by the reach of essential health services than by national wealth alone. Analysis shows that a 1% increase in DPT immunization coverage is associated with a 1.08% decline in mortality. This finding is consistent with the substantial impacts identified by McGovern and Canning (2015) and Li et al. (2021), who highlight that immunization remains one of the most effective tools for all-cause mortality reduction. Similarly, the 0.93% reduction linked to improved WASH access sits within the 0.7% to 1.2% range established by Cheng et al. (2012). These results indicate that environmental infrastructure and medical reach are the most reliable predictors of survival across all income tiers.2. The Efficiency Gap and Institutional MaturityWhile health spending is a foundational requirement, the model confirms that its effectiveness varies significantly by development level. The data indicates that spending is nearly twice as effective in high-income settings (-0.30) as it is in low-income settings (-0.16). This gap is supported by the global elasticity ranges identified in the meta-analysis by Gallet and Doucouliagos (2017). The reduced efficiency in lower-income nations supports the theory that institutional maturity acts as a multiplier for financial investment, as argued by Houtven and Norton (2020). Without stable supply chains and trained personnel—components of health system stability—increased budgets cannot achieve their full life-saving potential.3. Health System Stability vs. Economic GrowthBecause income status serves as a baseline indicator of health system stability, its loss of statistical significance (p > 0.05) when WASH and DPT are included suggests that economic classification is primarily a proxy for the quality of public services. Health system stability represents the underlying capacity of a country to maintain service delivery during economic or social shifts. As noted by Filmer and Pritchett (1999) and Bokhari et al. (2007), these findings suggest that low-income countries can outperform their wealth category and bypass the health outcomes of wealthier peers. They achieve this by prioritizing equitable access to vaccines and sanitation, effectively "leapfrogging" traditional development paths rather than relying solely on aggregate economic growth.

**Limitations of the findings**
While the methodology is rigorous, the study is subject to several important limitations that must be acknowledged for a balanced interpretation. First, the potential for residual confounding exists, as the model cannot account for all unobserved factors such as political stability, the quality of institutional governance, or maternal education levels, which likely influence both health investment and child survival. Second, while Quantile Regression with bootstrapping was utilized to mitigate data noise and heteroskedasticity, the inherent reporting gaps and potential measurement errors in low-income country data may still affect the precision of the estimated Efficiency Gap. Finally, as this is an observational study of 122 countries, these findings represent strong statistical associations rather than absolute causal claims, particularly since the relationship between spending and mortality can be bi-directional in high-burden settings.

```{r}

library(plm)      # Panel diagnostic tests
library(quantreg) # Quantile analysis
library(car)      # VIF/Diagnostics
library(tidyverse)# Data manipulation

# 1. PRELIMINARY: Income-Group Table
income_table <- data_final %>%
  group_by(income_status) %>%
  summarise(across(c(mortality_r, health_exp_perc_ppp, wash_access, dpt_coverage), 
                   ~mean(.x, na.rm = TRUE)))

# 2. LOG TRANSFORMATION
data_final$ln_mort <- log(data_final$mortality_r)
data_final$ln_spend <- log(data_final$health_exp_perc_ppp)

# 3. SELECTION TESTS (OLS-BASED)
# pdata setup
p_data <- pdata.frame(data_final, index = c("country", "year"))
formula_main <- ln_mort ~ ln_spend * income_status + wash_access + dpt_coverage

pooled <- plm(formula_main, data = p_data, model = "pooling")
fixed  <- plm(formula_main, data = p_data, model = "within")
random <- plm(formula_main, data = p_data, model = "random")

# THE HAUSMAN TEST: Deciding the model structure
phtest(fixed, random) 

# 4. FINAL ESTIMATION: Quantile Regression
# We analyze 10th (Mature), 50th (Median), and 90th (High Mortality) quantiles
final_model <- rq(ln_mort ~ ln_spend * income_status + wash_access + dpt_coverage, 
                  tau = c(0.1, 0.5, 0.9), data = p_data)

# Summary with Bootstrapped SE (handles 22-year panel autocorrelation)
summary(final_model, se = "boot")
```

