
---
title: "Economic Growth and Health System Factors in Global Child Mortality"
author: "Collins Audi"
date: "2025-12-04"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**BACKGROUND AND PURPOSE OF THIS HEALTH DATA ANALYSIS PROJECT**

Reducing the number of children who die before their fifth birthday is a core target within the sustainable development goal three of health and well-being and is a sign of a country’s human development index. Previously, the Preston curve showed a strong correlation between a country’s per capita income and under-5 deaths (Preston, 1975): whereby higher income levels are associated with lower under-five mortality.  Globally, the under-5 mortality has reduced by more than half from 94 per 1000 live births in 1990 to 37 per 1000 live births in 2023 (World Bank, 2023). This reduction can be attributed to economic growth and global health programs. 

Research in early 2000s established that economic growth alone is not enough to protect children unless it is paired with strategic priority-based investments in healthcare infrastructure. Houweling et al. (2005) reported that countries with comparable income levels experienced different outcomes because economic growth increased within-country inequalities and suggested increased public spending on health services like immunization and skilled birth attendance focusing on poor communities to close the gap. These inequalities were also confirmed by Cesar G Victoria (2003) who observed that the disparity between the rich and poor countries was becoming wider due to increasing income disparities within countries, and suggested increased interventions targeting the poor and equitable universal health coverage. 

Conversely, present evidence suggests that preventing mortality has evolved from access to the quality of care. Magret Kruk et al (2018) argues that despite the increased domestic and donor funding of health programs which has saved lives, the quality of care across different countries is often inadequate and recommends Universal Healthcare Coverage as an opportunity to improve the quality in health systems. The WHO Monitoring Report 2023 reported that while service coverage access improved across income status between 2015 and 2019, it was accompanied by limitations.According to this report, the service coverage index for healthcare has stagnated since the sustainable development goals were launched in 2015.The report further reveals how the focus of many countries has been on vertical programs rather than investing in health system capacities. In the report, while the low income and low middle income countries experienced catastrophic out-of-pocket expenditures despite service access gains, the upper middle income countries had gaps in access to services. This suggests that expenditure on healthcare by a country does not guarantee positive outcomes because it depends on the structural stability of the systems, yet this is not well quantified and documented. The report also revealed gaps in access to reproductive, maternal, neonatal, child and adolescent health services globally.While the importance of system pillars and spending is known, it remains unclear whether the efficiency of financial investment is conditional on the pre-existing structural maturity of the health system.

This project investigates these determinants of child survival through a quantitative analysis of cross national data, guided by two core research questions:

**Question**: To what extent do national income, government health spending, and preventive/infrastructural factors explain variations in child mortality, and does the effect of health spending vary by a country's income status, used as a proxy for structural health system maturity?

While high income status of countries may not automatically guarantee better access to healthcare as indicated above, it does have better health system stability at baseline. Hence, income status is utilized in this study as a proxy for health system stability and horizontal infrastructure required to for sustainable survival outcomes.
   
**Variables per country (2000-2002)**
    •	Under_5 mortality rate 
    •	Access to water and sanitation 
    •	Healthcare investment (as a % of GDP)
    •	GDP Per Capita (adjusted by Our World in Data to reflect living costs and inflation)
    •	Under-5 population size

  **Sources of Data**
     •	Our World in Data
     •	UNICEF
     •	World Bank
     
**STEP 1: IMPORTING, CLEANING AND MANIPULATING THE FIRST DATASET: GLOBAL UNDER_5 MORTALITY DATA**

**Step 1-Phase 1: Preparing the Mortality Data**
My first task was to take the raw mortality records from UNICEF and transform them into a format that works for statistical analysis. I followed these four steps:

**i. Importing the Data**: I loaded the UNICEF under 5 mortality file and performed an initial clean to remove any hidden spaces in the text that could cause errors later.

**ii. Selecting the Best Estimates**: The raw file contains three different numbers for every country: a high, low, and median estimate. I filtered the data to keep only the Median rows. This ensures my analysis is based on the most likely figures while removing unnecessary noise.

***iii. Cleaning Column Names***: When R imports data, it often adds an "X" in front of years (like X2000). I stripped these "X" prefixes away so the year columns would be clean and easy to read.

***iv. Reshaping for Analysis (The "Pivot")***: Most importantly, I changed the data from a "wide" layout to a "long" layout. Instead of having 30 separate columns for different years, I collapsed them into two columns: Year and Mortality Rate.

**v. The Result**: By the end of this step, I had a clean, vertical timeline for every country. This consistent structure is what allows me to merge the mortality data with other factors like wealth and health spending for the next part of my story.

```{r}
#Importing and managing the first data set by UNICEF: Under 5 mortality data
under_5_mortality <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/under_five_mortality.csv", header = TRUE, sep =',')
under_5_mortality

library(dplyr)
library(tidyr)
library(stringr)

colnames(under_5_mortality) <- str_trim(colnames(under_5_mortality))   # remove leading/trailing spaces

under_5_mortality <- under_5_mortality[, 1:37] %>%
  filter(bound == "Median") %>%   # keep only Median rows
  rename_with(~ gsub("^X", "", .x), 5:37)   # remove 'X' prefix from columns 5 to 37

# Step 1: Keeping only Median rows necessary, leaving upper and lower boundaries since our interest is in modelling.

under_5_mortality

colnames(under_5_mortality)

# Step 2: Pivot year columns into long format
under_5_mortality <- under_5_mortality %>%
  pivot_longer(
    cols = 5:37,
    names_to = "year",
    values_to = "mortality_r",
  ) %>%
  mutate(year        =  as.integer(year),
         mortality_r =  round(mortality_r, 1))

under_5_mortality
# Step 3: View the result


```
**Step 1-Phase 2: Importing the Health and Economic Indicators**

To build a complete picture of the factors influencing under 5 mortality, I imported and processed five additional datasets. I followed these steps for each one:

**i. Importing the Key Variables**: I loaded datasets for Vaccine Coverage (DPT shots), Healthcare Expenditure (as a percentage of GDP), GDP per capita, WASH Access (clean water and sanitation), and the total Under 5 Population.

**ii. Standardizing the Values**: To ensure the data was precise but easy to read, I applied a rounding function to all numeric columns. I rounded the percentages for vaccines and WASH, as well as the dollar amounts for GDP, to one decimal place.

**iii. Data Alignment**: Each of these files was checked to ensure they contained the necessary country and year information. This allows these specific health and economic indicators to be matched perfectly with the mortality data I prepared in Phase 1.

**The Result**: By the end of this phase, I had a collection of clean, standardized tables. Each table represents a different "piece of the puzzle"—from a country's total wealth and population size to its specific investments in health and infrastructure. These are now ready to be combined into a single master dataset.
```{r}
#Importing and managing the immunization data set
immunisation <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/immunisation.csv", header = TRUE, sep =',')

#Rounding off immunization coverage data to one decimal place
immunisation  <- immunisation %>%
  mutate(dpt_coverage = round(dpt_coverage, 1))

immunisation

#Importing and managing the healthcare expenditure data set as a % of GDP
h_expenditure <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/health_exp.csv", header = TRUE, sep =',')

#Rounding off health expenditure per capita figures to one decimal place
h_expenditure <- h_expenditure %>%
  mutate(health_exp_perc_ppp = round(health_exp_perc_ppp, 1))

#Importing gdp per capita data from Our World in Data
gdp_per_capita <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/gdp.csv", header = TRUE, sep =',')
gdp_per_capita

#Rounding off gdp per capita to one decimal place
gdp_per_capita <- gdp_per_capita %>%
  mutate(gdp_percapita = round(gdp_percapita, 1))

#Importing and managing the access to WASH (access to clean water) data set
wash <-read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/wash.csv", header = TRUE, sep =',')

wash <- wash %>%
  mutate(
    wash_access = round(wash_access, 1))
wash

#Importing Under 5 population data set
under5_pop <- read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/population.csv", header = TRUE, sep =',')

under5_pop

#Importing country income status data set
income <- read.csv("C:/Users/audic/Documents/GitHub/Beyond-GDP-Health-System-Drivers-of-Child-Survival/income_country.csv", header = TRUE, sep =',')

income
```
**Step 1-Phase 3: Data Merging and Coverage Audit**

In this final processing phase, I combined all individual datasets into one master file and conducted a thorough audit to check the quality of our timeline across different countries.

**Merging into a Master Dataset**: Using a "left join" function, I connected the mortality data with the five other datasets (Vaccines, WASH, GDP, Population, and Health Spending). I used the unique Country Code and Year as the keys to ensure every piece of information aligned perfectly across all sources.

**Cleaning the Combined Table**: After merging, the dataset contained several duplicate country name columns. I removed these extra columns and filtered the data to ensure that every row included the three most important variables for our bubble chart in the next exercise: GDP per capita, Mortality Rate, and Under 5 Population.

**Auditing the Timeline**: I conducted an audit to see how consistent the timeline was across different nations. I compared every country against the expected 23 year span (2000 to 2022).

**Full Range Coverage**: The audit revealed that 177 countries have a complete, uninterrupted record for the entire 23 year period.

**Partial Coverage**: Only 9 countries were identified as having partial or missing years. I kept these in the dataset to maintain global diversity, as reporting gaps are common in fragile or developing states.

**Final Export**: Once the data was verified, I exported the final collection as "merged_data.csv." This master file now contains 186 countries in total, providing a robust mix of stable and developing nations.

**The Result**: I now have a high quality master dataset. The fact that 177 countries have a full data range means that the patterns we see in the bubble chart are backed by a very large and reliable body of global evidence. This structure allows us to see not just where countries are today, but how they have moved over time.
```{r}
#Merging the datasets into a master dataset using iso_code and year columns, enabled by the left_join function
merged_df <- under_5_mortality %>%
  left_join(immunisation,  by = c("iso_code", "year")) %>%
  left_join(wash,          by = c("iso_code", "year")) %>%
  left_join(gdp_per_capita,by = c("iso_code", "year")) %>%
  left_join(under5_pop,    by = c("iso_code", "year")) %>%
  left_join(h_expenditure, by = c("iso_code", "year")) 


merged_df <- merged_df %>%
  select(
    -country.y,
    -country.y.y,
    -country.y.y.y,
    -country.x.x,
    -country.x.x.x,
    -bound,
     country = country.x) %>% 
  filter(!is.na(gdp_percapita), 
         !is.na(mortality_r), 
         !is.na(pop_under_5))

sapply(merged_df, class)

write.csv(merged_df, "merged_data.csv", row.names = FALSE)

#Now, we want to see the number of countries which have different year ranges 2000-2022 and 2001-2022
#Step 1: Counting the years available per country
country_years <- merged_df %>%
  group_by(iso_code, country) %>% # 1️⃣ Group the data by country
  summarise(
    min_year = min(year),         # 2️⃣ Earliest year for that country
    max_year = max(year),         # 3️⃣ Latest year for that country
    n_years  = n_distinct(year)   # 4️⃣ Count of unique years
  ) %>%
  ungroup() 
#group_by(iso_code, country.x) → ensures calculations are done separately for each country.
#min(year) → finds the earliest year recorded for that country.
#max(year) → finds the latest year recorded.
#n_distinct(year) → counts how many unique years exist for that country.
#ungroup() → removes grouping so we can do further operations easily.

#Step 2: Defining the expected full year range
expected_years <- 2000:2022           # 1️⃣ Full range of years
n_expected <- length(expected_years)  # 2️⃣ How many years: 23

#Step 3: Checking which countries have full 2000–2022 coverage
country_years <- country_years %>%
  mutate(
    full_2000_2022 = (min_year == 2000 & max_year == 2022 & n_years == n_expected))

#mutate() → adds a new column
#full_2000_2022 → TRUE if all conditions met:
#Earliest year = 2000
#Latest year = 2022
#Total number of years = 23
#FALSE → country is missing some years

#Step 4: Counting how many countries have full or partial coverage
country_years %>%
  summarise(
    full_range = sum(full_2000_2022),         # 1️⃣ Count TRUEs → complete countries
    partial_or_missing = sum(!full_2000_2022) # 2️⃣ Count FALSEs → incomplete countries
  )
```
*PHASE 2: VISUALIZING THE RELATIONSHIP BETWEEN UNDER-5 MORTALITY AND GDP**

As a preliminary step for answering the first question: **To what extent is a country’s GDP per capita associated with its under-five mortality rate (per 1000 live births) and does the link get weaker as countries get richer?**

The goal of this phase was to create an interactive "Preston Curve" (bubble chart) to visually explore the link between economic growth and child survival from 2000 to 2022.

**Dynamic Motion**: I built an animated bubble chart using Plotly, allowing us to see how countries move across the economic spectrum over time.

**Dimensions of Data**: Each bubble represents a country, with its position showing GDP per capita (x-axis) and Mortality Rate (y-axis). The size of each bubble is weighted by the Under-5 Population, highlighting where the largest groups of vulnerable children are located.Regional 

**Context**: I applied a custom color palette to group countries by their SDG Regions, making it easy to identify geographic trends and outliers.

**Logarithmic Scaling**: Because the differences between the richest and poorest nations are so vast, I used log scales for both axes. This ensures that progress in low-income countries is just as visible as trends in high-income ones.

**Preliminary Findings**: In this preliminary visualization, the bubble chart hints at a significant negative correlation between a country’s wealth and its child survival outcomes, suggesting that higher GDP per capita is generally linked to lower under-five mortality rates. This relationship follows a clear logarithmic pattern known as the Preston Curve: in lower-income countries, small increases in wealth are associated with declines in mortality, whereas in wealthier nations, the link becomes significantly weaker as survival rates approach the minimum possible death rate. For example, in 2000, Bangladesh (GDP - $2,712.9) had a high mortality rate of 85.4 per 1,000, but by 2022, as its wealth grew, it saw a dramatic drop in deaths to 31 deaths per 1000 live births. Conversely, in high-income regions like Europe and Northern America, bubbles are clustered at the bottom-right of the chart, showing that even substantial further increases in GDP result in only marginal improvements to already low mortality rates. Furthermore, the varying bubble sizes representing the under-five population highlight that high-population, lower-income nations like India and Bangladesh carry a disproportionate share of the global mortality burden despite their steady economic progress

```{r}
# ==========================================
# 1. LOAD LIBRARIES
# ==========================================
library(dplyr)    
library(ggplot2)  
library(plotly)   
library(scales)   

# ==========================================
# 2. DATA CLEANING & PRE-PROCESSING
# ==========================================
merged_df$country <- iconv(merged_df$country, from = "", to = "UTF-8", sub = "")
merged_df$country <- trimws(merged_df$country)
merged_df$country <- as.character(merged_df$country)

merged_df <- merged_df %>%
  filter(!is.na(mortality_r), country != "", !is.na(iso_code))

anim_df <- merged_df %>% 
  filter(year >= 2000, year <= 2022) %>% 
  group_by(iso_code, year) %>%          
  slice(1) %>%                           
  ungroup() %>%                          
  arrange(iso_code, year)               

# Strictly 1K and 100K for GDP; 1 and 500 for Mortality
x_breaks <- c(1000, 100000)  
y_breaks <- c(1, 500)       

# ==========================================
# 3. GGPLOT BASE CONSTRUCTION
# ==========================================
# Harmonized SDG region colors (soft, dark-background-friendly)
sdg_colors <- c(
  "Europe and Northern America"      = "#11A7B1", # Cyan/Teal
  "Northern Africa and Western Asia" = "#8E7AE3", # Purple
  "Central and Southern Asia"        = "#FF66C4", # Pink
  "Eastern and South-Eastern Asia"   = "#00B4FF", # Light Blue
  "Latin America and the Caribbean"  = "#4ECDC4", # Turquoise
  "Oceania"                          = "#FFA07A", # Light Salmon
  "Sub-Saharan Africa"               = "#FF7F7F"  # Light Red
)

p <- ggplot(
  anim_df,
  aes(
    x = gdp_percapita,       
    y = mortality_r,         
    color = sdg_region,       # Color by SDG region
    size = pop_under_5,      
    frame = year,            
    ids = iso_code,          
    group = iso_code,        
    text = paste(            
      "<b>", country, "</b>",
      "<br>Region:", sdg_region,
      "<br>Year:", year,
      "<br>GDP per Capita:", dollar(gdp_percapita),
      "<br>Under-5 Mortality:", round(mortality_r, 1),
      "<br>Under-5 Population:", comma(pop_under_5)
    )
  )
) +
  geom_point(alpha = 0.65) +  # Slightly transparent to reduce shouting

  scale_x_log10(
    limits = c(700, 150000),         
    breaks = x_breaks,               
    labels = dollar,                 
    expand = expansion(mult = 0.05)  
  ) +

  scale_y_log10(
    limits = c(1, 500),              
    breaks = y_breaks,               
    labels = comma,                  
    expand = expansion(mult = 0.05)
  ) +

  scale_size(range = c(4, 25), guide = "none") +
  scale_color_manual(values = sdg_colors, name = "SDG Region") +

  labs(
    x = "GDP per capita (USD, PPP Adjusted)",
    y = "Under-5 mortality (per 1,000 live births)"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "black", color = NA),  
    panel.background = element_rect(fill = "black", color = NA),
    text = element_text(color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.margin = margin(25, 25, 25, 25),
    axis.line = element_line(color = "white", linewidth = 0.6),
    axis.ticks = element_line(color = "white"),
    legend.position = "right",
    legend.background = element_rect(fill = "black", color = NA),
    legend.key = element_rect(fill = "black", color = NA),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white")
  )

# ==========================================
# 4. PLOTLY CONVERSION & INTERACTIVITY
# ==========================================
interactive_anim <- ggplotly(p, tooltip = "text") %>%
  animation_opts(
    frame = 4000,      
    transition = 4000, 
    easing = "cubic-in-out", 
    redraw = FALSE     
  ) %>%
  layout(
    xaxis = list(
      title = list(text = "GDP per capita (USD)", standoff = 25, font = list(size = 14, color = "white")),
      tickvals = log10(x_breaks), 
      ticktext = dollar(x_breaks),
      tickfont = list(size = 11, color = "white"),
      showgrid = FALSE
    ),
    yaxis = list(
      title = list(text = "Under-5 mortality (per 1,000 live births)", standoff = 15, font = list(size = 14, color = "white")),
      tickvals = log10(y_breaks),
      ticktext = comma(y_breaks),
      tickfont = list(size = 11, color = "white"),
      showgrid = FALSE
    ),
    title = list(
      text = paste0(
        "<span style='font-size:18px; color:#11A7B1;'><b>Higher GDP levels</b></span>",
        "<span style='font-size:18px; color:#CCCCCC;'>, as seen in wealthier countries,</span><br>",
        "<span style='font-size:18px; color:#CCCCCC;'>are generally associated with </span>",
        "<span style='font-size:18px; color:#8E7AE3;'><b>lower</b></span> ",
        "<span style='font-size:18px; color:#FF7F7F;'><b>child mortality rates</b></span>"

      ),
      x = 0.5, xanchor = "center", y = 0.88, yanchor = "top",
      font = list(family = "Arial, sans-serif")
    ),
    annotations = list(
      list(
        x = 0.88, y = 0.96, xref = "paper", yref = "paper",
        text = paste0(
          "<span style='color:#87CEEB'>Bubble size = Under-5 population</span>",
          "<span style='color:#cccccc'> &nbsp;&nbsp;|&nbsp;&nbsp; </span>",
          "<span style='color:#FFA07A'>Each bubble = A country (hover for details)</span>"
        ),
        showarrow = FALSE,
        font = list(size = 12, family = "Arial, sans-serif")
      )
    ),
    margin = list(t = 130, b = 80, l = 100, r = 40)
  )

# Display interactive plot
interactive_anim

# ==========================================
# 5. EXPORT
# ==========================================
htmlwidgets::saveWidget(
  interactive_anim,
  "under5_mortality_final.html",
  selfcontained = TRUE 
)

```

**Step 2-Phase 2: Building the Interactive Research Dashboard**

To make the findings more accessible to my audience, I developed an R Shiny Dashboard. While the core bubble chart remains the focus, the dashboard adds new layers of functionality that allow users to explore the data for themselves.

**Custom Filters**: I added a sidebar that allows users to filter the data by SDG Region or specific Countries. This is useful for looking at the "Wealth Health" story within a specific area, such as focusing only on Latin America or Sub Saharan Africa.

**Dynamic Comparison**: By selecting or deselecting countries, users can directly compare how different nations have moved along the Preston Curve. This makes it easy to see which countries are overperforming or underperforming compared to their neighbors.

**User Controlled Timeline**: The dashboard gives the user full control over the animation speed and the year slider. This allows for a deep dive into specific periods, such as the progress made during the era of the Millennium Development Goals.

**The Result**: This dashboard turns the data into an interactive experience. It allows anyone to test our research questions by looking at specific regions or countries.
 
```{r}
# Load required libraries for the Shiny application
library(shiny)    # Framework for building interactive web applications
library(dplyr)    # Data manipulation and transformation
library(ggplot2)  # Grammar of graphics for creating static plots
library(plotly)   # Converts ggplot2 plots to interactive web graphics
library(scales)   # Provides label formatting functions (dollar, comma, etc.)

# ==========================================
# DATA CLEANING AND PREPARATION SECTION
# ==========================================

# Convert country names to UTF-8 encoding to handle special characters
# iconv() converts character vectors between encodings
# 'from = ""' uses the current locale's encoding
# 'to = "UTF-8"' specifies target encoding
# 'sub = ""' replaces unconvertible characters with empty string
merged_df$country <- iconv(merged_df$country, from = "", to = "UTF-8", sub = "")

# Remove leading/trailing whitespace from country names
# trimws() eliminates spaces, tabs, newlines at the beginning/end of strings
merged_df$country <- trimws(merged_df$country)

# Convert country column to character type (ensures proper string handling)
# as.character() explicitly converts to string data type
merged_df$country <- as.character(merged_df$country)

# Filter the dataset to remove incomplete/irrelevant records
# %>% is the pipe operator from dplyr, passes left side result as first argument to right side
merged_df <- merged_df %>%
  filter(!is.na(mortality_r),  # Keep only rows where mortality_r is not missing (NA)
         country != "",         # Exclude empty string country names
         !is.na(iso_code))      # Exclude rows missing ISO country codes

# Create animation-ready dataset for years 2000-2022
anim_df <- merged_df %>%
  filter(year >= 2000, year <= 2022) %>%  # Limit to 21st century data range
  group_by(iso_code, year) %>%           # Group by unique country-year combinations
  slice(1) %>%                           # Keep only first row per group (removes duplicates)
  ungroup() %>%                          # Remove grouping structure for subsequent operations
  arrange(iso_code, year)                # Sort by country code then year for chronological order

# Define axis break points for logarithmic scales
# X-axis: GDP per capita breaks at $1,000 and $100,000
x_breaks <- c(1000, 100000)  # Creates numeric vector with two break points
# Y-axis: Mortality rate breaks at 1 and 500 deaths per 1,000
y_breaks <- c(1, 500)        # Creates numeric vector with two break points

# Extract unique SDG regions for dropdown menu
# unique() returns distinct values from the sdg_region column
all_regions <- unique(anim_df$sdg_region)
# Remove NA (missing) values from the region list
all_regions <- all_regions[!is.na(all_regions)]
# Sort regions alphabetically for consistent dropdown ordering
all_regions <- sort(all_regions)

# ==========================================
# USER INTERFACE (UI) DEFINITION
# ==========================================

# Create the main UI container using fluidPage (responsive layout)
ui <- fluidPage(
  
  # Custom CSS styling embedded in HTML head section
  tags$head(
    tags$style(HTML("
      /* Overall page styling - matches plotly dark theme */
      body {
        background-color: black !important;  /* Black background for entire page */
        color: white !important;             /* White text color throughout */
        font-family: 'Arial', sans-serif;    /* Consistent font family */
      }
      
      /* Main container removes default padding */
      .container-fluid {
        padding: 0 !important;               /* Remove all padding */
        background-color: black !important;  /* Ensure black background */
      }
      
      /* Sidebar panel styling */
      .well {
        background-color: #111111 !important;  /* Dark gray background */
        border: 1px solid #333333 !important;  /* Medium gray border */
        border-radius: 8px !important;         /* Rounded corners */
        padding: 20px !important;              /* Internal spacing */
        margin: 15px !important;               /* External spacing */
        box-shadow: 0 4px 12px rgba(0, 0, 0, 0.3); /* Subtle shadow for depth */
      }
      
      /* Sidebar header section */
      .sidebar-header {
        background-color: #1a1a2e;           /* Dark blue header background */
        color: white;                        /* White text */
        padding: 12px 15px;                  /* Vertical and horizontal padding */
        margin: -20px -20px 20px -20px;      /* Negative margins to extend to edges */
        border-radius: 8px 8px 0 0;          /* Rounded top corners only */
        font-weight: bold;                   /* Bold text */
        font-size: 15px;                     /* Slightly larger font */
        letter-spacing: 0.5px;               /* Slight spacing between letters */
        text-transform: uppercase;           /* All capital letters */
      }
      
      /* Dropdown input box styling */
      .selectize-input {
        background-color: #222222 !important;  /* Very dark gray background */
        color: white !important;               /* White text */
        border: 1px solid #444444 !important;  /* Light gray border */
        border-radius: 4px !important;         /* Slightly rounded corners */
        padding: 10px !important;              /* Internal padding for text */
      }
      
      /* Dropdown menu styling */
      .selectize-dropdown {
        background-color: #222222 !important;  /* Match input background */
        border = 1px solid #444444 !important; /* Consistent border */
      }
      
      /* Individual dropdown options */
      .selectize-dropdown-content div {
        background-color: #222222 !important;  /* Dark background for options */
        color: white !important;               /* White option text */
        padding: 8px 10px;                     /* Spacing within each option */
      }
      
      /* Hover effect for dropdown options */
      .selectize-dropdown-content div:hover {
        background-color: #2d3748 !important;  /* Lighter gray on hover */
      }
      
      /* Selected item in dropdown display */
      .selectize-input .item {
        color: #87CEEB !important;  /* Light blue color for selected items */
      }
      
      /* Form control labels */
      .control-label {
        color: #cccccc !important;  /* Light gray text */
        font-weight: 600;           /* Semi-bold weight */
        margin-bottom: 8px;         /* Space below label */
        display: block;             /* Block display for proper spacing */
        font-size: 14px;            /* Slightly smaller than body text */
      }
      
      /* Horizontal rule styling */
      hr {
        border-top: 1px solid #444444 !important;  /* Medium gray line */
        margin: 25px 0 !important;                 /* Space above and below */
      }
      
      /* Information box styling */
      .info-box {
        background-color: #1a1a2e;  /* Dark blue background */
        padding: 15px;              /* Internal spacing */
        border-radius: 6px;         /* Rounded corners */
        margin-top: 20px;           /* Space above the box */
      }
      
      /* Main plot container */
      .plot-container {
        padding: 10px;                /* Internal padding around plot */
        border-radius: 8px;           /* Rounded corners */
        background-color: #111111;    /* Dark gray background */
        margin: 15px;                 /* External spacing */
      }
      
      /* Error message styling */
      .shiny-output-error {
        color: #ff6b6b !important;    /* Red color for errors */
        font-weight: bold;            /* Bold text for emphasis */
        font-size: 16px;              /* Larger font size */
        padding: 20px;                /* Spacing around error */
      }
      
      /* Dropdown text input styling */
      .selectize-input input {
        color: white !important;      /* Ensure typed text is visible */
      }
      
      /* Focus state for dropdowns */
      .selectize-input.focus {
        border-color: #3498db !important;       /* Blue border when focused */
        box-shadow: 0 0 0 2px rgba(52, 152, 219, 0.2) !important; /* Blue glow */
      }
      
      /* Data sources footer */
      .data-sources {
        margin-top: 20px;             /* Space above footer */
        padding-top: 15px;            /* Space inside top */
        border-top: 1px solid #444444; /* Top border line */
        font-size: 12px;              /* Small font size */
        color: #999999;               /* Dimmed gray text */
      }
      
      /* Data sources title */
      .data-sources-title {
        color: #cccccc;               /* Light gray title */
        font-weight: 600;             /* Semi-bold */
        margin-bottom: 5px;           /* Space below title */
        font-size: 13px;              /* Slightly larger than content */
      }
    "))
  ),
  
  # ==========================================
  # LAYOUT STRUCTURE USING sidebarLayout
  # ==========================================
  
  # sidebarLayout creates two-panel layout: sidebar (controls) + main (plot)
  sidebarLayout(
    
    # ==========================================
    # SIDEBAR PANEL (Left side - Controls)
    # ==========================================
    sidebarPanel(
      width = 3,  # Occupies 3/12 of horizontal space (25% width)
      
      # Custom header for sidebar section
      div(class = "sidebar-header", 
          "Filter Controls"  # Section title
      ),
      
      # Region selector dropdown
      div(
        # Label for region dropdown
        tags$label("Select Region:", class = "control-label"),
        # Dropdown input for region selection
        selectInput(
          "region",                     # Input ID (referenced in server)
          label = NULL,                 # No visible label (handled by tags$label above)
          choices = c("All Regions", all_regions),  # Options: "All" plus unique regions
          selected = "All Regions",     # Default selection
          width = "100%"                # Full width of container
        )
      ),
      
      br(),  # Line break for spacing
      
      # Country selector dropdown
      div(
        # Label for country dropdown
        tags$label("Select Country:", class = "control-label"),
        # Dropdown input for country selection
        selectInput(
          "country",                     # Input ID (referenced in server)
          label = NULL,                  # No visible label
          choices = c("All Countries", sort(unique(anim_df$country))),  # All countries plus "All"
          selected = "All Countries",    # Default selection
          width = "100%"                 # Full width
        )
      ),
      
      hr(),  # Horizontal rule for visual separation
      
      # How-to-use instructions section
      div(
        # Section title with blue color
        h5("How to Use:", style = "color: #87CEEB; margin-top: 0;"),
        # Unordered list of instructions
        tags$ul(
          style = "color: #cccccc; font-size: 13px; padding-left: 20px;",  # List styling
          tags$li("Select a region or country to filter data"),      # Instruction 1
          tags$li("Hover over bubbles for country details"),         # Instruction 2
          tags$li("Use the slider below the chart to animate through years")  # Instruction 3
        )
      ),
      
      # Data sources attribution footer
      div(class = "data-sources",
          div(class = "data-sources-title", "Data Sources:"),  # Section title
          tags$div(style = "line-height: 1.4;",                # Content with line spacing
                   "• Our World in Data",                       # Source 1
                   tags$br(),                                   # Line break
                   "• UNICEF"                                   # Source 2
          )
      )
    ),
    
    # ==========================================
    # MAIN PANEL (Right side - Plot)
    # ==========================================
    mainPanel(
      width = 9,  # Occupies 9/12 of horizontal space (75% width)
      
      # Plot container div
      div(class = "plot-container",
          # Plotly output element - where the interactive chart will render
          plotlyOutput("bubble_plot",    # Output ID (matches server output$bubble_plot)
                       height = "600px", # Fixed height for rectangular aspect ratio
                       width = "100%")   # Responsive width (fills container)
      )
    )
  )
)

# ==========================================
# SERVER LOGIC DEFINITION
# ==========================================

# Server function defines reactive behavior and outputs
server <- function(input, output, session) {
  
  # ==========================================
  # SDG REGION COLOR DEFINITION
  # ==========================================
  
  # Color palette for SDG regions (soft colors for dark background)
  # Named vector: region names as keys, hex color codes as values
  sdg_colors <- c(
    "Europe and Northern America"      = "#11A7B1",  # Cyan/Teal
    "Northern Africa and Western Asia" = "#8E7AE3",  # Purple
    "Central and Southern Asia"        = "#FF66C4",  # Pink
    "Eastern and South-Eastern Asia"   = "#00B4FF",  # Light Blue
    "Latin America and the Caribbean"  = "#4ECDC4",  # Turquoise
    "Oceania"                          = "#FFA07A",  # Light Salmon
    "Sub-Saharan Africa"               = "#FF7F7F"   # Light Red
  )
  
  # ==========================================
  # REACTIVE OBSERVER: Update country dropdown
  # ==========================================
  
  # observeEvent triggers when input$region changes
  observeEvent(input$region, {
    
    # Determine which countries to show based on selected region
    if (input$region == "All Regions") {
      # If "All Regions" selected, show all countries from dataset
      country_choices <- sort(unique(anim_df$country))
    } else {
      # Filter countries to only those in selected region
      country_choices <- anim_df %>%
        filter(sdg_region == input$region) %>%  # Filter by selected region
        pull(country) %>%                       # Extract country column as vector
        unique() %>%                            # Get distinct country names
        sort()                                  # Sort alphabetically
    }
    
    # Update the country dropdown with new choices
    updateSelectInput(
      session,                       # Current Shiny session
      "country",                     # ID of input to update
      choices = c("All Countries", country_choices),  # New choices list
      selected = "All Countries"     # Reset to default selection
    )
  })
  
  # ==========================================
  # REACTIVE OUTPUT: Create bubble plot
  # ==========================================
  
  # renderPlotly creates the interactive plot output
  output$bubble_plot <- renderPlotly({
    
    # Start with full animation dataset
    filtered_data <- anim_df
    
    # Apply region filter if not "All Regions"
    if (input$region != "All Regions") {
      filtered_data <- filtered_data %>%
        filter(sdg_region == input$region)  # Filter to selected region
    }
    
    # Apply country filter if not "All Countries"
    if (input$country != "All Countries") {
      filtered_data <- filtered_data %>%
        filter(country == input$country)    # Filter to selected country
    }
    
    # Data validation - check if any data remains after filtering
    # validate() stops execution and shows error message if condition fails
    validate(
      need(nrow(filtered_data) > 0,         # Condition: must have at least 1 row
           paste("No data available for the selected filters.",
                 "Please try different region/country combinations."))  # Error message
    )
    
    # ==========================================
    # GGPLOT2 OBJECT CREATION
    # ==========================================
    
    # Create ggplot object with aesthetic mappings
    p <- ggplot(
      filtered_data,  # Data frame to use for plotting
      aes(
        x = gdp_percapita,       # X-axis: GDP per capita (economic indicator)
        y = mortality_r,         # Y-axis: Under-5 mortality rate (health indicator)
        color = sdg_region,      # Color encoding: SDG region (categorical)
        size = pop_under_5,      # Size encoding: Under-5 population (continuous)
        frame = year,            # Animation frame: Year (creates time dimension)
        ids = iso_code,          # Unique identifier: ISO country code
        group = iso_code,        # Grouping for animation: Country trajectories
        # Custom hover text using HTML formatting
        text = paste(            
          "<b>", country, "</b>",                     # Bold country name
          "<br>Region:", sdg_region,                  # Region information
          "<br>Year:", year,                          # Current year
          "<br>GDP per Capita:", dollar(gdp_percapita),  # Formatted GDP
          "<br>Under-5 Mortality:", round(mortality_r, 1),  # Rounded mortality rate
          "<br>Under-5 Population:", comma(pop_under_5)    # Formatted population
        )
      )
    ) +
      # Add points (bubbles) to the plot
      geom_point(alpha = 0.65) +  # Alpha = transparency (0.65 = 65% opaque)
      
      # ==========================================
      # X-AXIS SCALING (Logarithmic)
      # ==========================================
      scale_x_log10(
        limits = c(700, 150000),        # Axis range: $700 to $150,000
        breaks = x_breaks,              # Break points at $1K and $100K
        labels = dollar,                # Format as currency with dollar() from scales
        expand = expansion(mult = 0.05) # Add 5% padding on both ends
      ) +
      
      # ==========================================
      # Y-AXIS SCALING (Logarithmic)
      # ==========================================
      scale_y_log10(
        limits = c(1, 500),             # Axis range: 1 to 500 deaths/1000
        breaks = y_breaks,              # Break points at 1 and 500
        labels = comma,                 # Format numbers with comma separators
        expand = expansion(mult = 0.05) # Add 5% padding on both ends
      ) +
      
      # ==========================================
      # SIZE AND COLOR SCALES
      # ==========================================
      scale_size(range = c(4, 25),      # Bubble size range: 4 to 25 units
                 guide = "none") +      # Hide size legend (not needed)
      scale_color_manual(values = sdg_colors,  # Use custom color palette
                         name = "SDG Region") + # Legend title
      
      # ==========================================
      # AXIS AND PLOT LABELS
      # ==========================================
      labs(
        x = "GDP per capita (USD, PPP Adjusted)",  # X-axis label with detail
        y = "Under-5 mortality (per 1,000 live births)"  # Y-axis label
      ) +
      
      # ==========================================
      # THEME CUSTOMIZATION
      # ==========================================
      theme_minimal(base_size = 13) +  # Start with minimal theme, base font size 13
      theme(
        plot.background  = element_rect(fill = "black", color = NA),  # Black plot background
        panel.background = element_rect(fill = "black", color = NA),  # Black panel background
        text = element_text(color = "white"),            # All text in white
        axis.text = element_text(color = "white"),       # Axis numbers in white
        axis.title = element_text(color = "white", face = "bold"),  # Bold axis titles
        axis.title.y = element_text(margin = margin(r = 15, l = 5), vjust = 1),  # Y-title spacing
        axis.title.x = element_text(margin = margin(t = 15, b = 5)),  # X-title spacing
        panel.grid.major = element_blank(),     # Remove major grid lines
        panel.grid.minor = element_blank(),     # Remove minor grid lines
        plot.margin = margin(25, 25, 25, 25),   # Plot margins: top, right, bottom, left
        axis.text.x = element_text(margin = margin(t = 5)),   # X-axis number spacing
        axis.text.y = element_text(margin = margin(r = 5, l = 5)),  # Y-axis number spacing
        axis.line = element_line(color = "white", linewidth = 0.6),  # White axis lines
        axis.ticks = element_line(color = "white"),  # White axis ticks
      )
    
    # ==========================================
    # CONVERT TO PLOTLY INTERACTIVE PLOT
    # ==========================================
    
    # Convert ggplot object to interactive plotly object
    interactive_anim <- ggplotly(p,           # ggplot object to convert
                                 tooltip = "text",  # Use 'text' aesthetic for hover info
                                 width = NULL,      # Auto width
                                 height = NULL) %>% # Auto height
      # Animation options
      animation_opts(
        frame = 4000,             # Duration of each frame in milliseconds (4 seconds)
        transition = 4000,        # Transition time between frames (4 seconds)
        easing = "cubic-in-out",  # Smooth acceleration/deceleration
        redraw = FALSE            # Don't redraw entire plot each frame (better performance)
      ) %>%
      # Layout customization for plotly
      layout(
        # X-axis configuration
        xaxis = list(
          title = list(
            text = "GDP per capita (USD, PPP Adjusted)",  # X-axis title
            standoff = 25,          # Distance between title and axis
            font = list(size = 14, color = "white")  # Title font properties
          ),
          tickvals = log10(x_breaks),    # Positions for ticks on log scale
          ticktext = dollar(x_breaks),   # Labels for ticks (formatted as currency)
          tickfont = list(size = 11, color = "white"),  # Tick label styling
          showgrid = FALSE               # Hide grid lines
        ),
        
        # Y-axis configuration
        yaxis = list(
          title = list(
            text = "Under-5 mortality (per 1,000 live births)",  # Y-axis title
            standoff = 15,          # Distance between title and axis
            font = list(size = 14, color = "white")  # Title font properties
          ),
          tickvals = log10(y_breaks),    # Positions for ticks on log scale
          ticktext = comma(y_breaks),    # Labels for ticks (with comma formatting)
          tickfont = list(size = 11, color = "white"),  # Tick label styling
          showgrid = FALSE               # Hide grid lines
        ),
        
        # ==========================================
        # MAIN TITLE (Data Story)
        # ==========================================
        title = list(
          text = paste0(
            # First line: Blue emphasis on "Higher GDP levels"
            "<span style='font-size:18px; color:#11A7B1;'><b>Higher GDP levels</b></span>",
            # Continuation: Gray explanatory text
            "<span style='font-size:18px; color:#CCCCCC;'>, as seen in wealthier countries,</span><br>",
            # Second line start: More explanatory text
            "<span style='font-size:18px; color:#CCCCCC;'>are generally associated with </span>",
            # Purple emphasis on "lower"
            "<span style='font-size:18px; color:#8E7AE3;'><b>lower</b></span> ",
            # Red emphasis on "child mortality rates"
            "<span style='font-size:18px; color:#FF7F7F;'><b>child mortality rates</b></span>"
          ),
          x = 0.5,           # Horizontal position (center)
          xanchor = "center", # Anchor point is center of text
          y = 0.88,          # Vertical position (88% from bottom)
          yanchor = "top",   # Anchor point is top of text
          font = list(family = "Arial, sans-serif")  # Font family
        ),
        
        # ==========================================
        # SUBTITLE ANNOTATION
        # ==========================================
        annotations = list(
          list(
            x = 0.75,       # Horizontal position (88% from left)
            y = 0.99,       # Vertical position (96% from bottom)
            xref = "paper", # Use paper coordinates (0-1 scale)
            yref = "paper", # Use paper coordinates
            # Annotation text with color coding
            text = paste0(
              "<span style='color:#87CEEB'>Bubble size = Under-5 population</span>",  # Blue: size explanation
              "<span style='color:#cccccc'> &nbsp;&nbsp;|&nbsp;&nbsp; </span>",       # Gray separator
              "<span style='color:#FFA07A'>Each bubble = A country (hover for details)</span>"  # Salmon: hover tip
            ),
            showarrow = FALSE,  # Don't show arrow pointing to something
            font = list(size = 12, family = "Arial, sans-serif")  # Font properties
          )
        ),
        
        # Plot margins in pixels
        margin = list(t = 130,  # Top margin (extra space for title)
                      b = 80,   # Bottom margin
                      l = 100,  # Left margin
                      r = 40),  # Right margin
        
        autosize = TRUE,        # Automatically adjust to container size
        showlegend = FALSE       # Display the SDG region legend
      )
    
    # Return the final interactive plot object to be rendered
    interactive_anim
  })
}

# ==========================================
# RUN THE SHINY APPLICATION
# ==========================================

# Combine UI and server to create and launch the Shiny app
shinyApp(ui = ui, server = server)
```

```{r}
library(tidyverse)

library(dplyr)

merged_df2 <- under_5_mortality %>%
  left_join(immunisation %>% select(-any_of("country")),   by = c("iso_code", "year")) %>%
  left_join(wash %>% select(-any_of("country")),           by = c("iso_code", "year")) %>%
  left_join(gdp_per_capita %>% select(-any_of("country")), by = c("iso_code", "year")) %>%
  left_join(under5_pop %>% select(-any_of("country")),     by = c("iso_code", "year")) %>%
  left_join(h_expenditure %>% select(-any_of("country")),  by = c("iso_code", "year")) %>%
  left_join(income %>% select(-any_of("country")),         by = c("iso_code", "year")) %>%
  # Drop bound if it exists
  select(-any_of("bound")) %>%
  # This removes a row ONLY if ALL THREE are missing at the same time
  filter(!(is.na(wash_access) & is.na(health_exp_perc_ppp) & is.na(gdp_percapita)))

# Convert country names to UTF-8 encoding to handle special characters
# iconv() converts character vectors between encodings
# 'from = ""' uses the current locale's encoding
# 'to = "UTF-8"' specifies target encoding
# 'sub = ""' replaces unconvertible characters with empty string
merged_df2$country <- iconv(merged_df2$country, from = "", to = "UTF-8", sub = "")

# Remove leading/trailing whitespace from country names
# trimws() eliminates spaces, tabs, newlines at the beginning/end of strings
merged_df2$country <- trimws(merged_df2$country)

# Convert country column to character type (ensures proper string handling)
# as.character() explicitly converts to string data type
merged_df2$country <- as.character(merged_df2$country)

# 1. Define your exact target variables
target_vars <- c("mortality_r", "dpt_coverage", "wash_access", 
                 "gdp_percapita", "pop_under_5", "health_exp_perc_ppp")

# 2. Generate the "Problem Only" Table
failing_variables_audit <- merged_df2 %>%
  # Group by country to check each one's data health
  group_by(iso_code, country) %>%
  summarise(
    across(all_of(target_vars), ~ mean(is.na(.)) * 100), # Calculate % missing
    .groups = "drop"
  ) %>%
  # Reshape to see each variable on its own row for every country
  pivot_longer(cols = all_of(target_vars), 
               names_to = "Indicator", 
               values_to = "Missing_Percentage") %>%
  # FILTER: Only show cases that are 50% or above
  filter(Missing_Percentage >= 50) %>%
  arrange(country, desc(Missing_Percentage))

# 3. View the Resulting List
# This will ONLY show you the countries and variables that "fail" the rule.
print(failing_variables_audit)

# 1. Identify countries that are 100% missing in any variable
# MICE cannot impute if there is zero data to start with
critical_failures <- failing_variables_audit %>%
  filter(Missing_Percentage == 100) %>%
  distinct(iso_code) %>%
  pull(iso_code)

# 2. Identify countries failing more than 2 variables
high_failure_countries <- failing_variables_audit %>%
  group_by(iso_code) %>%
  count() %>%
  filter(n > 2) %>%
  pull(iso_code)
high_failure_countries

# 3. Combine and remove
final_drop_list <- unique(c(critical_failures, high_failure_countries))

merged_df3 <- merged_df2 %>%
  filter(
    !iso_code %in% final_drop_list,
    year >= 2000 & year <= 2022
  )


#Checking the structure of the remaining data
cat("=== TIER 1 DATA (All Countries) ===\n")
cat("Dimensions:", dim(merged_df3), "\n")
cat("Countries:", n_distinct(merged_df3$country), "\n")
cat("Years range:", min(merged_df3$year), "-", max(merged_df3$year), "\n")
cat("Variables:", names(merged_df3), "\n")
```
**Checking Missingness Patterns**
Missingness pattern was checked in details using heat maps.

```{r}
library(VIM)

aggr(merged_df3, prop = FALSE, combined = TRUE, numbers = TRUE, 
     cex.axis = 0.7, varheight = FALSE, sortVars = TRUE)

# Create a 'Missingness Flag' for the heatmap
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Reshape the data to long format
long_missing <- merged_df3 %>%
  select(country, year, wash_access, health_exp_perc_ppp, gdp_percapita, dpt_coverage) %>%
  pivot_longer(cols = c(wash_access, health_exp_perc_ppp, gdp_percapita, dpt_coverage), 
               names_to = "variable", 
               values_to = "value") %>%
  mutate(is_missing = is.na(value))

# 2. Create the Heatmap
# We use 'facet_grid' to separate variables while keeping countries on the Y-axis
p <- ggplot(long_missing, aes(x = year, y = country, fill = is_missing)) +
  geom_tile(color = "white") +
  facet_wrap(~variable) +  # This creates 3 side-by-side heatmaps
  scale_fill_manual(values = c("grey90", "red"), 
                    labels = c("Data Present", "Missing (NA)")) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7),
        strip.text = element_text(face = "bold")) + 
  labs(title = "Detailed Missingness: Country vs. Year vs. Variable",
       x = "Year", y = "Country")
p
# 3. Save as a TALL image to ensure names are readable
ggsave("Detailed_Variable_Heatmap.png", plot = p, width = 15, height = 25, limitsize = FALSE)
```

**Checking if missingness is dependent on other variables**

Visual analysis revealed systematic missingness patterns across key variables:

**1. WASH Indicators:**

- Countries with missing WASH data show mortality rate distributions shifted toward substantially higher values (extending toward 300-500 range)
- Countries with observed WASH data show mortality concentrated at lower values (0-100 range)
**Pattern:** Missingness strongly associated with worse health outcomes

**2. Health Expenditure:**
- Different population distributions between observed and missing groups
- Pattern: Missingness relates to country characteristics, not random

**3. GDP Data:**
- Missing GDP associated with higher mortality distributions
- Missing GDP concentrated in earlier years (2000-2010)
**Pattern:** Dual association with both poor outcomes and historical reporting gaps

**Missingness Mechanism Assessment**
**MCAR (Missing Completely at Random):** Ruled out due to systematic patterns.
**MAR (Missing at Random):** Plausible if key predictors are included:
**mortality_r** for WASH/GDP missingness
**pop_under_5** for health expenditure missingness year for GDP missingness

**MNAR (Missing Not at Random):** Probable for WASH/GDP given extreme mortality associations.

**Imputation Justification**
MICE is conditionally justified because:
- Missingness patterns are explainable by observed variables
- Including key predictors makes MAR assumption tenable
```{r}
library(mice)
library(lattice) # For vignette-style histograms
library(dplyr)

# Create indicators: TRUE = Missing, FALSE = Observed
merged_df3 <- merged_df3 %>%
  mutate(
    M_wash   = is.na(wash_access),
    M_gdp    = is.na(gdp_percapita),
    M_health = is.na(health_exp_perc_ppp),
    M_dpt    = is.na(dpt_coverage)
  )

# A. Check if Mortality predicts missingness in WASH (Target 1)
p1 <- histogram(~ mortality_r | M_wash, data = merged_df3,
          main = "Mortality Rate by WASH Missingness",
          strip = strip.custom(factor.levels = c("WASH Obs", "WASH Miss")))

# B. Check if Mortality predicts missingness in GDP (Target 2)
p2 <- histogram(~ mortality_r | M_gdp, data = merged_df3,
          main = "Mortality Rate by GDP Missingness",
          strip = strip.custom(factor.levels = c("GDP Obs", "GDP Miss")))

# C. Check if Pop Under 5 predicts missingness in Health Exp (Target 3)
p3 <- histogram(~ pop_under_5 | M_health, data = merged_df3,
          main = "Pop by Health Exp Missingness",
          strip = strip.custom(factor.levels = c("Health Obs", "Health Miss")))

p4 <- histogram(~ year | M_wash, data = merged_df3,
                main = "Year by WASH Missingness",
                strip = strip.custom(factor.levels = c("WASH Obs", "WASH Miss")))

library(gridExtra)

grid.arrange(p1, p2, p3, p4, ncol = 2)
```
**Initial Diagnosis** Missing data was identified as Missing At Random (MAR) because missingness patterns depended on observed variables like Year and Mortality Rate.

**Problem 1:** Impossible Values & Poor Shape: The initial model (2l.pan) produced impossible negative values and failed to replicate the "humps" in the data distribution.

**Change:** Switched to 2l.pmm (Predictive Mean Matching) from the miceadds package.

**Reason:** PMM ensures no negative results by "borrowing" real observed values, which preserves the natural bimodal shape of your data.

**Problem 2:** "Optimistic" Bias: Density plots showed that imputed values for wash_access were too high, missing the lower-access (0–50) hump entirely.

**Change:** Added sdg_region and income_status to the predictor matrix.

**Reason:** These variables provided the necessary economic and geographic "clues" for the model to correctly identify and match missing data to low-access donors.

**Problem 3:** Formula Syntax Errors: The code crashed with "unexpected symbol" errors.

**Change:** Used gsub to replace spaces/hyphens with underscores and converted strings to factors.

**Reason:** R multilevel formulas cannot process special characters or raw strings within factor levels.

**Problem 4:** Numerical Spikes & Scale Warnings: GDP data produced "spiky" distributions (over-smoothing) and scaling warnings due to its high numerical range.

**Change:** Increased iterations to maxit = 30 and integrated categorical anchors.

**Reason:** This stabilized the model, allowing for a diverse spread of imputed values that matched the original distribution despite the large scale.

**Final Result:** Validated the success via stripplots and density plots, confirming that all imputed values are now bounded, realistic, and capture the true bimodal nature of the variables.
```{r}
# ============================================================================
# REVISED MICE PIPELINE (UNSCALED GDP + INCOME STATUS)
# ============================================================================

# 1. LOAD LIBRARIES ----------------------------------------------------------
library(mice)
library(miceadds)  # Essential for 2l.pmm
library(dplyr)
library(lattice)

# 2. PREPARE DATA ------------------------------------------------------------
df_prepared <- merged_df3 %>%
  filter(!(is.na(wash_access) & is.na(dpt_coverage) & is.na(gdp_percapita))) %>%
  mutate(
    iso_code = as.integer(factor(iso_code)), 
    # CLEAN CATEGORICALS: Critical for formula compatibility
    sdg_region = as.factor(gsub("[[:space:]|-]", "_", as.character(sdg_region))),
    income_status = as.factor(gsub("[[:space:]|-]", "_", as.character(income_status))),
    year = as.numeric(year),
    time_in_country = year - 2000
  )

# 3. INITIALIZE MICE ---------------------------------------------------------
init <- mice(df_prepared, maxit = 0, printFlag = FALSE)
meth <- init$method
pred <- init$predictorMatrix

# 4. IMPUTATION METHODS ------------------------------------------------------
# 2l.pmm prevents negative predictions naturally
meth["wash_access"] <- "2l.pmm"
meth["gdp_percapita"] <- "2l.pmm"
meth["health_exp_perc_ppp"] <- "2l.pmm"
meth["dpt_coverage"] <- ""        
meth["sdg_region"] <- "polyreg"
meth["income_status"] <- "polyreg" 

# 5. PREDICTOR MATRIX --------------------------------------------------------
pred <- make.predictorMatrix(df_prepared)
pred[, "country"] <- 0

# Using raw gdp_percapita as requested
essential_preds <- c("mortality_r", "pop_under_5", "time_in_country", 
                     "sdg_region", "income_status", "gdp_percapita")

for (var in c("wash_access", "gdp_percapita", "health_exp_perc_ppp")) {
  pred[var, ] <- 0
  pred[var, essential_preds] <- 1
  pred[var, "iso_code"] <- -2     # Set iso_code as the Level-2 cluster
  pred[var, "dpt_coverage"] <- 1
}

# 6. RUN IMPUTATION ----------------------------------------------------------
set.seed(123)
imp_complete <- mice(
  df_prepared,
  m = 10,
  method = meth,
  predictorMatrix = pred,
  maxit = 30, # Increased iterations for better convergence with large scales
  printFlag = TRUE
)

# 7. VALIDATION -------------------------------------------------------------
# Note: Step 9 (pmax/pmin) is deleted. PMM handles bounds.
densityplot(
  imp_complete,
  ~ wash_access + gdp_percapita + health_exp_perc_ppp,
  layout = c(3, 1),
  main = "Final Validation: Observed vs Imputed"
)

# 8. CONVERGENCE (TRACE) PLOTS -----------------------------------------------
# This creates plots for the mean and SD of each variable across iterations
plot(imp_complete, 
     c("wash_access", "gdp_percapita", "health_exp_perc_ppp"),
     main = "Convergence Check: Mean and SD per Iteration")

# 9. STRIP PLOTS (Another way to see Imputation numbers) ----------------------
# This plots the raw imputed values side-by-side for each imputation dataset
stripplot(
  imp_complete, 
  wash_access + gdp_percapita + health_exp_perc_ppp ~ .imp, 
  pch = 20, 
  cex = 0.5,
  main = "Individual Imputed Values by Imputation Number"
)

# View the first imputed dataset (m = 1)
data_final <- complete(imp_complete, 1)
write.csv(data_imputed_1, "final_imputed.csv")
# 10. SAVE OUTPUT ------------------------------------------------------------
saveRDS(imp_complete, "imputed_data_unscaled.rds")
```
**Regression Analysis of the Final Panel Data with Imputed Values**
**Analysis Process and Model Selection**
The study utilizes a 22-year panel dataset of 122 countries to examine the factors that determine under-five mortality. The analysis focuses on the relationship between child survival (ln_mort: the natural log of the child mortality rate, used to measure percentage changes rather than raw unit drops) and its primary drivers: financial investment (ln_spend: the natural log of health spending per person in PPP dollars, representing the total volume of resources), infrastructure access (wash_access: the percentage of the population with basic water and sanitation services), and preventive healthcare (dpt_coverage: the percentage of children receiving the DPT vaccine as a measure of how well primary clinics reach the population). Income Status (World Bank income groups) is used as the proxy for structural maturity, representing the permanent foundations of a health system, such as trained staff and stable supply chains.

All money and mortality variables were log-transformed to measure elasticities, which allow the results to show the percentage change in survival for every 1% increase in spending. While Quantile Regression is naturally robust to skewed data, these transformations were still applied to stabilize variance across the 122 countries and to linearize the relationship between spending and health, ensuring mathematical accuracy. To ensure a rigorous model selection, a three-stage econometric hierarchy was followed. First, an F-test confirmed that country-specific characteristics must be accounted for. Second, the Hausman Test yielded a highly significant p-value (6.472e-06), rejecting the Random Effects model in favor of Fixed Effects. Finally, Panel Quantile Regression with bootstrapped standard errors was selected as the main analytical tool. This approach accounts for heteroscedasticity (uneven noise in the data) by re-sampling the data thousands of times to provide reliable, "criticism-proof" p-values.

**Key Findings**
Foundations of Survival and Systemic Pillars The results confirm that saving children's lives depends on both funding and the underlying strength of the system. A 10% increase in health spending (ln_spend) is linked to a 4.65% reduction in child mortality (ln_mort) for the baseline group. Furthermore, basic infrastructure and preventive care show universal success across the entire dataset: every 1% increase in WASH access and vaccination (DPT) coverage leads to a mortality drop of 1.17% and 0.97%, respectively. These findings prove that while money is essential, it must be supported by a solid foundation of basic services to be effective.

The Efficiency Gap and Structural Maturity The analysis provides clear evidence of an Efficiency Gap (the difference in how much "result" a country gets from its health dollars) based on its level of structural maturity. In countries with the highest death rates (the 90th quantile), health spending in Low-Income settings is far less effective; a positive interaction value (ln_spend : income_status) of +0.442 almost completely cancels out the benefit of the spending. This suggests that in systems with low structural maturity, health funding often hits "bottlenecks" that prevent it from saving lives. However, as countries improve and mortality rates fall (the 10th quantile), this gap disappears (-0.001), showing that health spending becomes significantly more efficient as the structural maturity of the health system increases.

**Limitations of the findings**
While the methodology is rigorous, the study is subject to several important limitations that must be acknowledged for a balanced interpretation. First, the potential for residual confounding exists, as the model cannot account for all unobserved factors such as political stability, the quality of institutional governance, or maternal education levels, which likely influence both health investment and child survival. Second, while Quantile Regression with bootstrapping was utilized to mitigate data noise and heteroskedasticity, the inherent reporting gaps and potential measurement errors in low-income country data may still affect the precision of the estimated Efficiency Gap. Finally, as this is an observational study of 122 countries, these findings represent strong statistical associations rather than absolute causal claims, particularly since the relationship between spending and mortality can be bi-directional in high-burden settings.

```{r}

library(plm)      # Panel diagnostic tests
library(quantreg) # Quantile analysis
library(car)      # VIF/Diagnostics
library(tidyverse)# Data manipulation

# 1. PRELIMINARY: Income-Group Table
income_table <- data_final %>%
  group_by(income_status) %>%
  summarise(across(c(mortality_r, health_exp_perc_ppp, wash_access, dpt_coverage), 
                   ~mean(.x, na.rm = TRUE)))

# 2. LOG TRANSFORMATION
data_final$ln_mort <- log(data_final$mortality_r)
data_final$ln_spend <- log(data_final$health_exp_perc_ppp)

# 3. SELECTION TESTS (OLS-BASED)
# pdata setup
p_data <- pdata.frame(data_final, index = c("country", "year"))
formula_main <- ln_mort ~ ln_spend * income_status + wash_access + dpt_coverage

pooled <- plm(formula_main, data = p_data, model = "pooling")
fixed  <- plm(formula_main, data = p_data, model = "within")
random <- plm(formula_main, data = p_data, model = "random")

# THE HAUSMAN TEST: Deciding the model structure
phtest(fixed, random) 

# 4. FINAL ESTIMATION: Quantile Regression
# We analyze 10th (Mature), 50th (Median), and 90th (High Mortality) quantiles
final_model <- rq(ln_mort ~ ln_spend * income_status + wash_access + dpt_coverage, 
                  tau = c(0.1, 0.5, 0.9), data = p_data)

# Summary with Bootstrapped SE (handles 22-year panel autocorrelation)
summary(final_model, se = "boot")
```


```{r}
# Check the correlation between your independent variables
# A high number (e.g., > 0.7) suggests Multicollinearity
cor(main_data_final[, c("wash_access", "gdp_percapita", "dpt_coverage", "health_exp_perc_gdp")])
```


```{r}
# Null Hypothesis: Pooled OLS is better
# Alternative: Fixed Effects is better
pFtest(fixed, pooled)
```


